---
title: "Fit Analysis"
output: html_document
---

For my first post I thought I'd write about some basic data classification. I've been going through the courses given by Carnegie Melon in the [Data Science Specialization Stream](https://www.coursera.org/specialization/jhudatascience/1) to learn R and some 
introductory data science techniques. I recommend it. 

Anyway, for the [Practical Machine Learning Course](https://class.coursera.org/predmachlearn-006) they asked us to classify the how well people were performing bicep curls. The data as well as the explanation of the problem is [here](http://groupware.les.inf.puc-rio.br/har). We have data taken from subjects wearing sensors while they perform a bicep curl with five differing techniques ("A", "B", "C", "D", "E"). "A" corresponds to a "proper" bicep curl and the rest are differing bad techniques. The goal is to use this data to construct a classifier that can then classify the bicep curl using only the sensory data.   

The [random forest](http://cran.r-project.org/web/packages/randomForest/index.html) worked extremely well for this. 

```{r, message = FALSE}
library(randomForest)
library(ggplot2)
```
Read in the training and testing data
```{r}
train = read.csv("pml-training.csv", stringsAsFactors = FALSE)
test = read.csv("pml-testing.csv", stringsAsFactors = FALSE)
```
I wrote a function to find the indices of the columns in the training data which do have not
enough data. They need to have more than 406 entries which are not blank or NA.
```{r}
#function to find index of columns with very little usable data
#from training data (blanks and NA's)
bad.cols <- function(dataframe, threshold = 406){
  
  #count how many non NA's are in a column
  count.non.nas =  function(x){
    sum(!is.na(x)) 
  }
  #find columns with too many na's
  na.cols = apply(dataframe, MARGIN = 2, count.non.nas) > threshold
  na.cols = which(na.cols == F)
  
  #count how many columns do not have blanks
  count.non.blanks = function(x){
    sum(!(nchar(x) == 0))
  }
  #remove columns with too many blanks
  blank.cols = apply(dataframe, MARGIN = 2, count.non.blanks) > threshold
  blank.cols = which(blank.cols == F)
  
  return(unique(c(na.cols, blank.cols)))
}
```
The indices of the unnecessary columns are 
```{r}
no.use.col = bad.cols(train)
no.use.col = c(1:7, no.use.col)
no.use.col
```
Now I prepare the data by separating out the classification column and removing the unnecessary columns. Since the classification column will not be used for analysis I remove it.
```{r}
train.class = as.factor(train$classe)

train = train[, -no.use.col]
train$classe = NULL

test = test[, -no.use.col]
test$problem_id = NULL
```
I did everything by hand and didn't use the caret package. Not the smartest way to do things. I am
using PCA to reduce dimensionality. First have to center and rescale the data. The centering and scaling comes from the training data. I use the same centering and rescaling for the test data.
```{r}
#calculate mean and sds for each column of train
means = colMeans(train)
sds = apply(train, 2, sd)

#center and rescale
train = as.data.frame(t((t(train) - means)/sds))
test = as.data.frame(t((t(test) - means)/sds))

#do PCA
train.prcomp = prcomp(train, center = FALSE)

#predict using PCA
train = as.data.frame(predict(train.prcomp, newdata = train))
test = as.data.frame(predict(train.prcomp, newdata = test))
```
A plot of the first two principal components of the training set. Do they separate 
the data well?
```{r}
qplot(PC1, PC2, data = train, colour = train.class)
```

I did some analysis to find how many PC's were necessary by splitting the train data into training and validation data. This is probably not a good idea since I calculated means and sd's for the full train set. Should have split the data up top. 
```{r}
# train.cut = train[, c(1:25)] #use only a few of PCA's
# train.cut$classe = train.class #add back in classification column
# 
# set.seed(1224)
# fit = randomForest(classe~., data=train.cut,
#                importance = TRUE, ntree = 130)
# predictions = predict(fit, newdata = test)
# predictions
```


