---
title: "WaterMap"
author: "Ben Rifkind"
date: "January 8, 2015"
output: html_document
---
```{r, echo=FALSE, message=FALSE}
library(knitr)
library(ggplot2)
```
This is just a quick writeup on how to use GIS files to visualize some data from [Open Data](http://www1.toronto.ca/wps/portal/contentonly?vgnextoid=9e56e03bb8d1e310VgnVCM10000071d60f89RCRD). The idea was to visualize the water consumption of the different wards of Toronto. I have a bit of an easy point to make. My guess is that that the richer you are the more water you consume. The water consumption of each ward is available [here](http://www1.toronto.ca/wps/portal/contentonly?vgnextoid=289bd103cd8b1310VgnVCM1000003dd60f89RCRD&vgnextchannel=8517e03bb8d1e310VgnVCM10000071d60f89RCRD).
The data is water consumption for each ward from 2000 to 2013. 

First I read in the data for 2000 and 2013.
```{r}
water.2000 = read.csv('CoT_Water_Consumption/Water_Consumption_2000.csv', header = T)
water.2013 = read.csv('CoT_Water_Consumption/Water_Consumption_2013.csv', header = T)
```


```{r, echo = FALSE}
#data clean up
clean.water = function(toronto.water){
  #remove last row which has the totals
  totals = tail(toronto.water, n = 1)
  toronto.water = head(toronto.water, n = -1)
  # fix the comma issue - ie., numbers such as 1,234.56
  fixCommas = function(x) { as.numeric(gsub(',', '', x)) }
  toronto.water = apply(toronto.water, c(1,2), fixCommas)
  return(as.data.frame(toronto.water))
  }
water.2000 = clean.water(water.2000)
water.2013 = clean.water(water.2013)
```


There is data for commercial accounts and residential accounts. The ammounts are all in metres cubed. Here's a quick look at the data for the first 3 wards.
```{r, results="asis", echo=FALSE}
kable(water.2000[1:3, 1:5])
kable(water.2000[1:3, c(1,2, 6:8)])
```


Here is a quick plot of average water consumption over the years. I faceted the data according to whether it was residential or commercial use.
```{r, echo = FALSE, message=FALSE, fig.width = 10}
library(reshape2)
water = rbind(water.2000, water.2013)
water = water[, c(1,2,5,8)] #subset out usage values 
melt.water = melt(water, id.vars = c("city.ward", "year"), 
             variable.name = "usage.type", value.name = "usage.ammount")
levels(melt.water$usage.type) = c("Average Residential", "Average Commercial")

p = ggplot(data = melt.water, aes(x=city.ward, y = usage.ammount)) +
  geom_path(aes(color=factor(year))) +
  scale_colour_discrete(name  ="Year") +
  facet_grid(usage.type~., scales = "free_y") +
  xlab("City Wards") +
  ylab("Average Usage") +
  labs(title = "Water Usage in metres cubed") + 
  scale_x_discrete(breaks=1:44, labels=as.character(1:44)) +
  theme(axis.text.x = element_text(size=6)) 
p
```


So the ammount of average water consumption looks like it has been going down over the years. This is a little unexpected and heartening for me. I was also interested in seeing how the water consumption varies throughout the different wards of Toronto. Who is using the most water? I thought a good visualization of this would be to construct a heat map of the water consumption. To do this I used the GIS information for ward boundaries available at [open data Toronto](http://www1.toronto.ca/wps/portal/contentonly?vgnextoid=b1533f0aacaaa210VgnVCM1000006cd60f89RCRD&vgnextchannel=1a66e03bb8d1e310VgnVCM10000071d60f89RCRD). 

First read in the shape file and rotate it so that the topmost boundary is more or less flat. 
```{r, message=FALSE, fig.width=8, fig.height=10}
library(maptools)
ward.map = readShapeSpatial('gcc/icitw_wgs84.shp',
                               proj4string = CRS("+proj=longlat +datum=WGS84") )
ward.map = elide(ward.map, rotate=13)
```

First thing I do is use fortify to turn the spatial object into a data frame. Then I use ggplot to map it and the polygons that are the ward boundaries.
```{r, echo = FALSE, message=FALSE}
# use this theme for my map plots
map.theme = theme(axis.text=element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          panel.border = element_blank())  
```

```{r, message=FALSE}
library(rgeos)
ward.map.df = fortify(ward.map, region = "SCODE_NAME")
ward.map.df$id = as.integer(ward.map.df$id) #fixes ward labels
ggplot(data=ward.map.df, aes(x=long, y=lat, group=group)) + 
  geom_polygon(fill = NA, colour = 'black') + 
  map.theme #theme for map plots not defined here
```


This looks very basic. I'll add the ward numbers to each ward and make that heat map. The first step is to merge the water data and map data together. I also find the centroids of each polygon and label them according to the ward. I will use this labelling to identify the wards in the final plot.
```{r, message=FALSE}
# merge the water data and map data together
total.df = merge(ward.map.df, water, by.x = 'id', by.y = 'city.ward')
total.df = total.df[order(total.df$order), ]

#find centroids of polygons for ward labelling
centroids  = as.data.frame(gCentroid(ward.map, byid = T))
# keep track of which polygon the centroid is in
centroids[['id']] = ward.map@data$SCODE_NAME
```


Now I can make that heat map. Here are the residential maps. (I didn't include the code for the 2013 map since it would be redundant.)
```{r heatMaps, fig.height=6, fig.width = 12.5}
total.df.2000 = total.df[total.df$year==2000,]
pAverageResidential <- ggplot(data = total.df.2000, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = average.residential.usage), color = 'red', alpha = 0.6) +
  scale_fill_gradient(name = 'Average residential\nusage in metres cubed',
                      low = "cyan3", high = "blue3") +
  labs(title = "Average Toronto Residential Water Usage in 2000")

pAverageResidential + 
  map.theme + #theme for a map plot I defined earlier
  geom_text(aes(x=x,y=y, group=NULL, label=id), data = centroids, size=2) #ward labels
```

```{r, echo=FALSE, fig.height=6, fig.width = 12.5, fig.cap = "Code not included"}
total.df.2013 = total.df[total.df$year==2013,]
pAverageResidential <- ggplot(data = total.df.2000, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = average.residential.usage), color = 'red', alpha = 0.6) +
  scale_fill_gradient(name = 'Average residential\nusage in metres cubed',
                      low = "cyan3", high = "blue3") +
  labs(title = "Average Toronto Residential Water Usage in 2013")

pAverageResidential + 
  map.theme + #theme for a map plot I defined earlier
  geom_text(aes(x=x,y=y, group=NULL, label=id), data = centroids, size=2) #ward labels
```

So water consumption throughout Toronto has gone down but still the downtown core definitely seems to be using the most water. And in particular, ward 20, Trinity-Spadina area of Torontom is using quite a bit more water per person than the rest of Toronto. I really don't have any good idea why this might be. I initially thought the wealthier areas of Toronto would consume more water. The simplest factor I had in mind was swimming pools. But it looks like this does not matter so much.


Here are the maps for commercial water consumption.
```{r, echo = FALSE, fig.height=6, fig.width = 12.5}
pAverageCommercial.2000 <- ggplot(data = total.df.2000, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = average.commercial.usage), color = 'red', alpha = 0.6) +
  scale_fill_gradient(name = 'Average commercial\nusage in metres cubed',
                      low = "cyan3", high = "blue3") +
  labs(title = "Average Toronto Commercial Water Usage in 2000")

pAverageCommercial.2013 <- ggplot(data = total.df.2013, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = average.commercial.usage), color = 'red', alpha = 0.6) +
  scale_fill_gradient(name = 'Average commercial\nusage in metres cubed',
                      low = "cyan3", high = "blue3") +
    labs(title = "Average Toronto Commercial Water Usage in 2013")

pAverageCommercial.2000 + 
  map.theme + 
  geom_text(aes(x=x,y=y, group=NULL, label=id), data = centroids, size=2)

pAverageCommercial.2000 + 
  map.theme + 
  geom_text(aes(x=x,y=y, group=NULL, label=id), data = centroids, size=2)

```
It looks like ward 2, Etobicoke, is the heavy user of water commercially. Any ideas why this would be the case?